---
title: "Lab 13"
author: "Iris Thesmar"
format:
  html:
    toc: true
    toc_float: true
    embed-resources: true
execute: 
  warning: false
  message: false
---

```{r}
library(tidyverse) 
library(lubridate)
library(DT)
library(viridis)
library(janitor)
library(plotly)
library(respirometry)
```

```{r}
NEON_soilMAGs_soilChem <- read.csv("NEON_soilMAGs_soilChem.csv")
```

```{r}
#install.packages("vegan")

#install.packages("corrplot")

# Load the package
library(corrplot)
library(vegan)   # for ecological analyses
library(sf)      # for spatial data
library(ggplot2)
```

```{r}
# Replace with your actual file path
soil_data <- read_csv("NEON_soilMAGs_soilChem.csv")

colnames(soil_data)
```

# Environmental correlations

Environmental correlations of the different type of micro organisms found in the soil. The goal is too see whether the organisms are more or less associated with a certain soil moisture, or soil temperature.

```{r}
# Environmental variables
env_vars <- soil_data %>%
  select(soilMoisture.x, soilMoisture.y, soilTemp,
         soilInWaterpH, soilInCaClpH,
         decimalLatitude, decimalLongitude, elevation)

# Genomic metadata (MAG features)
mag_vars <- soil_data %>%
  select(bin_completeness, bin_contamination,
         average_coverage, gene_count, scaffold_count)

```

## Correlation test on Ph and communities

We want to see the the relationship between the multiple variables through a heat map.

```{r}
#| fig-height: 6
#| fig-width: 15
library(fastDummies)              # run every session

mag_vars <- soil_data %>%
  select(soilInWaterpH, soilInCaClpH, phylum,
        community, bin_contamination)

# Dummy-encode phylum + community
mag_numeric <- dummy_cols(mag_vars, 
                          select_columns = c("phylum", "community"),
                          remove_first_dummy = TRUE,   # avoids collinearity
                          remove_selected_columns = TRUE)

clean_data <- na.omit(soil_data)

cor_matrix <- cor(mag_numeric, use = "pairwise.complete.obs")
# Heatmap option
heatmap(cor_matrix, scale = "none")
```

Rather inconclusive, as one can admire.

```{r}
#| fig-height: 8
#| fig-width: 10
# Now rerun your correlation plot
corrplot(cor_matrix, method = "color", tl.cex = 0.9, cl.cex = 0.9)
```

## Ordination analysis to observe the correlation between features and environmental variables

```{r}
combined <- bind_cols(mag_vars, env_vars) %>%
  filter(if_all(everything(), ~ !is.na(.) & is.finite(.)))

mag_clean <- combined %>% select(bin_completeness, bin_contamination,
                                 gene_count, scaffold_count)
env_clean <- combined %>% select(soilMoisture.x, soilMoisture.y, soilTemp,
                                 soilInWaterpH, soilInCaClpH,
                                 decimalLatitude, decimalLongitude, elevation)

rda_model <- rda(mag_clean ~ ., data = env_clean)
```

```{r}
#| fig-height: 6
#| fig-width: 10

# Base RDA plot with arrows
plot(rda_model, main = "RDA: MAG features vs Environmental Variables")


```

```{r}
library(ggrepel)              # every session
# --- Extract scores ---
sp  <- scores(rda_model, display = "species", scaling = 2)  # species/MAG features
st  <- scores(rda_model, display = "sites",   scaling = 2)  # samples
bp  <- scores(rda_model, display = "bp",      scaling = 2)  # environmental arrows

sp_df <- as.data.frame(sp)
sp_df$label <- rownames(sp_df)

st_df <- as.data.frame(st)
st_df$label <- rownames(st_df)

bp_df <- as.data.frame(bp)
bp_df$label <- rownames(bp_df)

# --- Variance explained percentages ---
var <- summary(rda_model)$cont$importance[2,1:2] * 100

# --- Build ggplot ---
ggplot() +
  
  # site points
  geom_point(data = st_df,
             aes(x = RDA1, y = RDA2),
             color = "black", alpha = 0.7, size = 2) +
  
  # species points
  geom_point(data = sp_df,
             aes(x = RDA1, y = RDA2),
             color = "red", size = 2) +
  
  # species labels (non-overlapping)
  geom_text_repel(data = sp_df,
                  aes(x = RDA1, y = RDA2, label = label),
                  color = "red", size = 3, max.overlaps = Inf) +
  
  # environmental arrows
  geom_segment(data = bp_df,
               aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
               arrow = arrow(length = unit(0.25,"cm")),
               color = "blue", linewidth = 0.7) +
  
  # environmental labels (non-overlapping)
  geom_text_repel(data = bp_df,
                  aes(x = RDA1, y = RDA2, label = label),
                  color = "blue", size = 3, max.overlaps = Inf) +
  
  # axis labels using explained variance
  xlab(paste0("RDA1 (", round(var[1], 1), "%)")) +
  ylab(paste0("RDA2 (", round(var[2], 1), "%)")) +
  
  theme_classic() +
  ggtitle("RDA: MAG features vs Environmental Variables")

```

# Geographical analysis

```{r}
taxonomy_split <- soil_data %>%
  separate(phylum, into = c("Domain","Phylum","Class","Order","Family","Genus"),
           sep = ";", fill = "right", remove = FALSE)
```

```{r}
comm_matrix <- taxonomy_split %>%
  group_by(site, phylum) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = phylum, values_from = n, values_fill = 0)
```

```{r}
comm_matrix <- comm_matrix %>% filter(!is.na(site))
comm_matrix$site <- make.unique(comm_matrix$site)


site_taxa <- comm_matrix %>% column_to_rownames("site")
site_taxa <- comm_matrix %>% column_to_rownames("site")
```

```{r}
site_taxa <- site_taxa %>% mutate(across(everything(), as.numeric))
site_taxa <- as.data.frame(site_taxa)
site_taxa[is.na(site_taxa)] <- 0
```

```{r}
#| fig-height: 10
#| fig-width: 15

#install.packages("ggrepel")
library(ggrepel)

nmds <- metaMDS(site_taxa, distance = "bray", k = 2, trymax = 100)

site_scores <- as.data.frame(scores(nmds, display = "sites"))
site_scores$SiteID <- rownames(site_scores)

taxa_scores <- as.data.frame(scores(nmds, display = "species"))
taxa_scores$Taxon <- rownames(taxa_scores)

ggplot() +
  geom_point(data = site_scores, aes(x = NMDS1, y = NMDS2), color = "black") +
  geom_text_repel(data = site_scores, aes(x = NMDS1, y = NMDS2, label = SiteID), color = "black") +
  geom_text_repel(data = taxa_scores, aes(x = NMDS1, y = NMDS2, label = Taxon), color = "red") +
  labs(title = "NMDS of Microbial Communities", x = "NMDS1", y = "NMDS2") +
  theme_minimal()
```

```{r}
# Community matrix
comm_matrix <- taxonomy_split %>%
  filter(!is.na(siteID), !is.na(phylum)) %>%
  group_by(siteID, phylum) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = phylum, values_from = n, values_fill = 0)

site_taxa <- comm_matrix %>%
  column_to_rownames("siteID") %>%
  as.data.frame()

# Collapse coords to unique sites
coords <- soil_data %>%
  distinct(siteID, decimalLatitude, decimalLongitude)

# Keep only sites present in site_taxa
common_sites <- intersect(rownames(site_taxa), coords$siteID)

coords <- coords %>%
  filter(siteID %in% common_sites)

# Reorder coords to match site_taxa
coords <- coords[match(common_sites, coords$siteID), ]

# Check alignment
all(coords$siteID == rownames(site_taxa))  # should return TRUE

comm_dist <- vegdist(site_taxa, method = "bray")
geo_dist  <- dist(coords %>% select(decimalLatitude, decimalLongitude))

mantel_result <- mantel(comm_dist, geo_dist)
print(mantel_result)
```

```{r}
# Convert distance objects to matrices
comm_mat <- as.matrix(comm_dist)
geo_mat  <- as.matrix(geo_dist)

# Extract lower triangle (unique site pairs only)
comm_vec <- comm_mat[lower.tri(comm_mat)]
geo_vec  <- geo_mat[lower.tri(geo_mat)]

# Build tidy data frame
df_decay <- data.frame(
  GeographicDistance = geo_vec,
  CommunitySimilarity = 1 - comm_vec
)

# Plot with ggplot2
ggplot(df_decay, aes(x = GeographicDistance, y = CommunitySimilarity)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(
    title = "Distance–Decay Relationship",
    x = "Geographic Distance",
    y = "Community Similarity (1 - Bray-Curtis)"
  ) +
  theme_minimal()
```

# Taxonomy analysis

```{r}
#| fig-height: 6
#| fig-width: 8
taxonomy_split <- soil_data %>%
  separate(gtdb_taxonomy_lineage, into = c("Domain","Phylum","Class","Order","Family","Genus"),
           sep = ";", fill = "right", remove = FALSE)

# --- Summarize composition at Phylum level ---
phylum_counts <- taxonomy_split %>%
  count(phylum, sort = TRUE) %>%
  mutate(RelativeAbundance = n / sum(n) * 100)

# --- Plot relative abundance ---
ggplot(phylum_counts, aes(x = reorder(phylum, -RelativeAbundance),
                          y = RelativeAbundance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Taxonomic Composition at Phylum Level",
       x = "Phylum",
       y = "Relative Abundance (%)") +
  theme_minimal()
```

```{r}
#| fig-height: 10
#| fig-width: 15
# --- Compare across locations (example) ---
# Replace 'Location' with the actual column name in your dataset
phylum_by_location <- taxonomy_split %>%
  group_by(site, Phylum) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(site) %>%
  mutate(RelativeAbundance = n / sum(n) * 100)

ggplot(phylum_by_location, aes(x = site, y = RelativeAbundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  labs(title = "Phylum Composition by Location",
       x = "Location",
       y = "Relative Abundance (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
